---
title: "Practical_01: API exercise"
output: github_document
date: "2026-04-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data
Data is collected using the FRED API. The GINI coefficient - an indicator of inequality - is obtained for Chile and Argentina from 1987 to 2024. Further, the USD prices of commodities (copper, zinc, and aluminum) per metric ton  are collected.


# Installation, data collection, and tidying
Install (if necessary) and load fredr, pacman, and tidyverse. 

```{r}
if(!require ( "pacman" , quietly = TRUE ) ) {
   install.packages("pacman")
   library(pacman)
   }
if(!require ( "fredr" , quietly = TRUE ) ) {
  install.packages("fredr")
  library(fredr)
  }
library(tidyverse)
```

# Data collection
Use the FRED API key and indicator names to source the time series. 

```{r}
# Set your FRED API key - obtain from https://fredaccount.stlouisfed.org/apikey
fredr_set_key("ec84f06f1b178b454f2923dc06f22591")

# Fetch Chile and Argentina Gini Coefficients

chile_gini <- fredr(series_id = "SIPOVGINICHL" , observation_start = as.Date("1987-01-01"))

argentina_gini <- fredr(series_id = "SIPOVGINIARG" , observation_start = as.Date("1987-01-01"))

# Fetch commodity prices

copper_price <- fredr(series_id = "PCOPPUSDM" , observation_start = as.Date("1992-01-01"))

aluminium_price <- fredr(series_id = "PALUMUSDM" , observation_start = as.Date("1992-01-01"))

zinc_price <- fredr(series_id = "PZINCUSDM" , observation_start = as.Date("1992-01-01"))
```


# Data cleaning and tidying

Columns meaningfully renamed, the Chile and Argentina dataframes are merged into an "Inequality" dataframe, and commodity prices are merged into a "Commodity Price" dataframe. The dataframes are then tidied by removing unnecessary columns and by using pivot_longer to make new "country" and "commodity" columns in the respective dataframes.

```{r}
# Tidy the data for the first plot

arg_gini <- argentina_gini %>% 
  rename(
    argentina = value
  )

chil_gini <- chile_gini %>% 
  rename(
    chile = value
  )

# Merge Chile and Argentina and select relevant columns
merge01 <- arg_gini %>% 
  left_join(
    chil_gini,
    by = "date"
  ) %>% 
  select(date, argentina, chile)

# Pivot longer
inequality <- merge01 %>% 
  pivot_longer(
    cols = argentina:chile,
     names_to = "country",
     values_to = "gini"
  )

# Tidy the data for Plot 2

# Rename "value" column to commodity
copper_price <- copper_price %>% 
  rename(
    copper_price = value
  )

aluminium_price <- aluminium_price %>% 
  rename(
    aluminium_price = value
  )

zinc_price <- zinc_price %>% 
  rename(
    zinc_price = value
  )

# Merge commodity prices

merge02 <- copper_price %>% 
  left_join(
    aluminium_price,
    by = "date"
  ) %>% 
  select(date, copper_price, aluminium_price)

merge03 <- merge02 %>% 
  left_join(
    zinc_price,
    by = "date"
  ) %>% 
  select(date, copper_price, aluminium_price, zinc_price)

# Pivot longer
commodity_prices <- merge03 %>% 
  pivot_longer(
    cols = copper_price:zinc_price,
    names_to = "commodity",
    values_to = "price"
  )
```


# Plots
ggplot is used to visualise the data from the cleaned and tidied dataframes.

## Plot 1: Chile and Argentina Gini Coefficients

```{r}
# Plot 1: Chile and Argentina Gini Coefficients
ggplot(inequality,
       aes(x = date, y = gini, colour = country)) +
  geom_point() +
  geom_smooth() +
  labs( title = "Chile and Argentina Gini Coeffecients", y = "Gini", x = "Year")
```

This plot explores the movement of Chile and Argentina's respective Gini coefficients since 1987. This annual figure indicates the level of inequality present in an economy. Chile a much higher level of inequality compared to Argentina, but over the years the two have seemingly converged. 
## Plot 2: Commodity Prices

```{r}
#Plot 2: Commodity Prices
ggplot(commodity_prices,
       aes(x = date, y = price, colour = commodity)) +
  geom_line() +
  labs( title = "Commodity Prices", y = "USD per Metric Ton", x = "Year")
```

The second plot explores the movements in monthly commodity prices since 1992. Until the mid-2000s, copper, zinc, and aluminium were in close range of each other. Copper has since diverged in price and around three times the price of aluminium and zinc in the age of the green transition and growing demand for copper.
